---
title: "ECMO US"
author: "Joshua Geiger"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
options(width = 2000)

library(tidyverse)
library(broom)
library(janitor)
library(readxl)
library(tableone)
library(ROCR)
library(stringr)
library(pROC)
library(gtsummary)
library(caret)

ecmo <- read_xlsx("raw_data/VascularECMOdataset_11_5_23_mrn_cleaned.xlsx",
                 na = c("", "NA", ".", "9999", "Not applicable")) %>%
  clean_names()

```

A total of `r length(unique(ecmo$e_mrn))` patients were identified since January 2013 with a total of `r nrow(ecmo)` ecmo cannulations. The data set was filtered for patients > 18 years old who had peripheral VA - ECMO (Femoral cannulation).

## Descriptive statistics

```{r data-cleaning, include=FALSE}

ecmo <- ecmo %>%
  mutate(sex = case_when(sex == "F" ~ "Female",
                         sex == "M" ~ "Male",
                         .default = sex),
####demographics          
         age = as.numeric(age),
        race_ethnicity = case_when(race_ethnicity == "Asian" ~ "Other",
                                   race_ethnicity == "Hispanic/Latino" ~ "Other",
                                   race_ethnicity ==  "Native American or Alaska Native" ~ "Other",
                                   .default = race_ethnicity),
        cannulation_location = case_when(cannulation_location == "Non-ICU WARD" ~ "Other",
                                         cannulation_location == "OSH Transfer, already on ECMO" ~ "OSH",
                                         cannulation_location == "OSH (transfer)" ~ "OSH",
                                         cannulation_location == "Team goes to OSH ‚Üí cannulates patient ‚Üí transfers to Primary Facility" ~ "OSH",
                                         .default = cannulation_location),
#### comorbidities
        chronic_lung_disease = case_when(past_medical_history_choice_chronic_lung_disease == 1 ~ "Yes", 
                                         past_medical_history_choice_chronic_lung_disease == 0 ~ "No",
                         .default = NA),
        DM = case_when(past_medical_history_choice_diabetes_mellitus == 1 ~ "Yes",
                       past_medical_history_choice_diabetes_mellitus == 0 ~ "No",
                       .default = NA),
        HLD = case_when(past_medical_history_choice_hyperlipidemia == 1 ~ "Yes",
                       past_medical_history_choice_hyperlipidemia == 0 ~ "No",
                       .default = NA),
        shock_after_cardiac_arrest = case_when(etiology_of_shock_choice_cardiac_arrest_code == 1 ~ "Yes",
                                etiology_of_shock_choice_cardiac_arrest_code == 0 ~ "No",
                                .default = NA),
        cardiogenic_shock = case_when(indication_for_ecmo_choice_cardiogenic_shock == 1 ~ "Yes",
                                      indication_for_ecmo_choice_mixed_shock == 1 ~ "Yes",
                                      shock_after_cardiac_arrest == "Yes" ~ "Yes",
                                      etiology_of_shock_choice_decompensated_ischemic_cardiomyopathy == 1 ~ "Yes",
                                      etiology_of_shock_choice_decompensated_nonischemic_cardiomyopathy == 1 ~ "Yes",
                                      etiology_of_shock_choice_heart_failure == 1 ~ "Yes",
                                      etiology_of_shock_choice_myocardial_infarction == 1 ~ "Yes",
                                      .default = "No"),
        smoking_tobacco_use = case_when(smoking_tobacco_use %in% c("None", "none") ~ "None",
                                        smoking_tobacco_use %in% c("remote", "Remote") ~ "Remote",
                                        str_detect(smoking_tobacco_use, "Recent") ~ "Recent",
                                        .default = NA),
        anticoagulation = case_when(type_of_anticoagulation_choice_bivalirudin_drip == 1 ~ "Yes",
                                    type_of_anticoagulation_choice_heparin_drip == 1 ~ "Yes",
                                    type_of_anticoagulation_choice_lovenox_injections == 1 ~ "Yes",
                                    type_of_anticoagulation_choice_bivalirudin_drip == 0 &
                                      type_of_anticoagulation_choice_heparin_drip == 0 &
                                      type_of_anticoagulation_choice_lovenox_injections == 0 ~ "No",
                                    .default = NA),
#### US velocities
         sfa_distal_psv = case_when(!is.na(sfa_distal_pvs_on_ecmo_pre_rpc) ~ sfa_distal_pvs_on_ecmo_pre_rpc,
                                    !is.na(sfa_distal_pvs_on_ecmo_post_rpc) ~ sfa_distal_pvs_on_ecmo_post_rpc,
                                    .default = NA),
         sfa_mid_psv = case_when(!is.na(sfa_mid_pvs_on_ecmo_pre_rpc) ~ sfa_mid_pvs_on_ecmo_pre_rpc,
                                    !is.na(sfa_mid_pvs_on_ecmo_post_rpc) ~ sfa_mid_pvs_on_ecmo_post_rpc,
                                 .default = NA),
         sfa_prox_psv = case_when(!is.na(sfa_prox_pvs_on_ecmo_pre_rpc) ~ sfa_prox_pvs_on_ecmo_pre_rpc,
                                    !is.na(sfa_prox_pvs_on_ecmo_post_rpc) ~ sfa_prox_pvs_on_ecmo_post_rpc,
                                  .default = NA),
         sfa_prox_mid_psv = case_when(!is.na(sfa_prox_pvs_on_ecmo_pre_rpc) ~ sfa_prox_pvs_on_ecmo_pre_rpc,
                             !is.na(sfa_mid_pvs_on_ecmo_pre_rpc) ~ sfa_mid_pvs_on_ecmo_pre_rpc,
                             !is.na(sfa_prox_pvs_on_ecmo_post_rpc) ~ sfa_prox_pvs_on_ecmo_post_rpc,
                             !is.na(sfa_mid_pvs_on_ecmo_post_rpc) ~ sfa_mid_pvs_on_ecmo_post_rpc,
                             .default = NA),
         ata_psv = case_when(!is.na(ata_pvs_on_ecmo_pre_rpc) ~ ata_pvs_on_ecmo_pre_rpc,
                                    !is.na(ata_pvs_on_ecmo_post_rpc) ~ ata_pvs_on_ecmo_post_rpc),
         pta_psv = case_when(!is.na(pta_pvs_on_ecmo_pre_rpc) ~ pta_pvs_on_ecmo_pre_rpc,
                                    !is.na(pta_pvs_on_ecmo_post_rpc) ~ pta_pvs_on_ecmo_post_rpc),
         peroneal_psv = case_when(!is.na(peroneal_pvs_on_ecmo_pre_rpc) ~ peroneal_pvs_on_ecmo_pre_rpc,
                                    !is.na(peroneal_pvs_on_ecmo_post_rpc) ~ peroneal_pvs_on_ecmo_post_rpc),
         peak_ankle_velocity = pmax(ata_psv, pta_psv, peroneal_psv, na.rm = TRUE),
         log_pav = log(peak_ankle_velocity +1),
         femoral_at_gradient = sfa_prox_mid_psv - ata_psv,
         femoral_pt_gradient = sfa_prox_mid_psv - pta_psv,
         max_femoral_pt_at_gradient = pmax(femoral_at_gradient, femoral_pt_gradient, na.rm = TRUE),
         contra_sfa_psv = case_when(!is.na(contra_sfa_prox_pvs_on_ecmo_pre_rpc) ~ contra_sfa_prox_pvs_on_ecmo_pre_rpc,
                                    !is.na(contra_sfa_prox_psv_on_ecmo_post_rpc) ~ contra_sfa_prox_psv_on_ecmo_post_rpc,
                                  .default = NA),
         preop_abi_l = case_when(preop_abi_l == "Non-compressible" ~ "2.00",
                                 .default = preop_abi_l),
         preop_abi_l = as.numeric(preop_abi_l),
         preop_abi_r = case_when(preop_abi_r == "Non-compressible" ~ "2.00",
                                 .default = preop_abi_r),
         preop_abi_r = as.numeric(preop_abi_r),
         preop_abi_r = case_when(preop_abi_r >2.0 ~ 2.0,
                                 .default = preop_abi_r),
         preop_abi_l = case_when(preop_abi_l >2.0 ~ 2.0,
                                 .default = preop_abi_l),
 
#### outcomes
         hospital_length_of_stay = as.numeric(hospital_length_of_stay),
         amputation_yn = case_when(amputation == 1 ~ "Yes",
                                   amputation == 0 ~ "No",
                                   .default = NA),
         fasciotomy_yn = case_when(fasciotomy == 1 ~ "Yes",
                                   fasciotomy == 0 ~ "No",
                                   .default = NA),
         mortality_yn = case_when(mortality == 1 ~ "Yes",
                                   mortality == 0 ~ "No",
                                   .default = NA),
#create variable of death while on ecmo
         ecmo_mortality_yn = case_when(str_detect(survival_status_mcs_hospitalization,
                                                  "Died on MCS") ~ 1,
                                       mortality_date <= ecmo_discontinuation_date_time ~ 1,
                                       str_detect(survival_status_mcs_hospitalization,
                                                  "explant") ~ 0,
                                       mortality_date > ecmo_discontinuation_date_time ~ 0,
                                       mortality == 0 ~ 0,
                                       .default = NA),
         limb_ischemia_yn = case_when(hypoperfusion_limb_ischemia == 1 ~ "Yes",
                                      fasciotomy == 1 ~ "Yes",
                                      hypoperfusion_limb_ischemia == 0 ~ "No",
                                      fasciotomy == 0 ~ "No",
                                     .default = NA),
         ischemic_comp_yn = case_when(amputation_yn == "Yes" | fasciotomy_yn == "Yes" | limb_ischemia_yn == "Yes" ~ 1,
                                     amputation_yn == "No" ~ 0,
                                     fasciotomy_yn == "No" ~ 0,
                                     limb_ischemia_yn == "No" ~ 0,
                                     .default = NA)) %>%
  ### more comorbidities
        mutate(pad_yn = case_when(pad == 1 ~ "Yes",
                           preop_abi_l < 0.9 ~ "Yes",
                           preop_abi_r < 0.9 ~ "Yes",
                           pad == 0 ~ "No",
                           preop_abi_l > 0.9 & preop_abi_l < 1.2 &
                             preop_abi_r > 0.9 & preop_abi_l < 1.2 ~ "No",
                           .default = NA),
          distal_perfusion_cannula_ever = case_when(distal_perfusion_cannula == "Yes" ~ "Yes",
                                                      distal_perfusion_cannula == "No" ~ "No",
                                                      .default = NA),
          same_side_inflow_outflow = case_when(outflow_cannulation_side == inflow_cannulation_side ~ "Yes",
                                               outflow_cannulation_side != inflow_cannulation_side ~ "No",
                                               .default = NA)) %>%
  select(!etiology_of_shock_choice_cardiac_arrest_code) %>%
  rename(rosc_prior_to_ecmo = return_of_spontaneous_circulation_rosc_prior_to_ecmo_deployment) %>%
  mutate(us = case_when(sfa_distal_pvs_on_ecmo_pre_rpc >= 0 ~ 1,
                        sfa_mid_pvs_on_ecmo_pre_rpc >= 0 ~ 1,
                        sfa_prox_pvs_on_ecmo_pre_rpc  >= 0 ~ 1,
                        pop_pvs_on_ecmo_pre_rpc >= 0 ~ 1,
                        peroneal_pvs_on_ecmo_pre_rpc >= 0 ~ 1,
                        ata_pvs_on_ecmo_pre_rpc >= 0 ~ 1,
                        pta_pvs_on_ecmo_pre_rpc >= 0 ~ 1,
                        sfa_prox_pvs_on_ecmo_post_rpc >= 0 ~ 1,
                        sfa_mid_pvs_on_ecmo_post_rpc >= 0 ~ 1,
                        sfa_distal_pvs_on_ecmo_post_rpc >= 0 ~ 1,
                        pop_pvs_on_ecmo_post_rpc >= 0 ~ 1,
                        peroneal_pvs_on_ecmo_post_rpc >= 0 ~ 1,
                        ata_pvs_on_ecmo_post_rpc >= 0 ~ 1,
                        pta_pvs_on_ecmo_post_rpc >= 0 ~ 1,
                          .default = 0),
         reperfusion_cath_at_us = case_when(!is.na(sfa_distal_pvs_on_ecmo_pre_rpc) ~ 1,
                        !is.na(sfa_mid_pvs_on_ecmo_pre_rpc) ~ 1,
                        !is.na(sfa_prox_pvs_on_ecmo_pre_rpc) ~ 1,
                        !is.na(pop_pvs_on_ecmo_pre_rpc) ~ 1,
                        !is.na(peroneal_pvs_on_ecmo_pre_rpc) ~ 1,
                        !is.na(ata_pvs_on_ecmo_pre_rpc) ~ 1,
                        !is.na(pta_pvs_on_ecmo_pre_rpc) ~ 1,
                          .default = 0),
         distal_perfusion_at_cannulation =
            case_when(distal_perfusion_cannula_done_at_the_time_of_initial_cannulation == "Yes" ~ "Yes",
                      distal_perfusion_cannula_done_at_the_time_of_initial_cannulation == "No" ~ "No",
                      reperfusion_cath_at_us == 0 ~ "No",
                      .default = NA)) %>%
### calculate means across columns for velocities  
  rowwise() %>%
  mutate(mean_sfa_psv = mean(c_across(c(sfa_prox_psv, sfa_mid_psv, sfa_distal_psv)), na.rm = TRUE),
         average_ankle_velocity = mean(c_across(c(ata_psv, pta_psv, peroneal_psv)), na.rm = TRUE)) 

### variable dependent on means
ecmo <- ecmo %>%
  mutate(contra_sfa_aav_gradiant = contra_sfa_psv - average_ankle_velocity)

### identify etiology of shock duplicate values
shock_duplicate <-  ecmo %>%
  select(c(starts_with("etiology_of_shock_choice"), study_record_id)) %>%
  pivot_longer(cols = starts_with("etiology_of_shock_choice"), names_to = "etiology_of_shock",
               names_prefix = "etiology_of_shock_choice_") %>%
  filter(value == 1) %>%
  group_by(study_record_id) %>%
  summarise(n = n()) %>%
  filter(n > 1)

# remove "other" duplicates
ecmo <- ecmo %>%
  mutate(etiology_of_shock_choice_other = 
           case_when(study_record_id %in% shock_duplicate$study_record_id ~ 0,
                     etiology_of_shock_choice_post_cardiotomy_shock == 1 ~ 1,
                     etiology_of_shock_choice_ards == 1 ~ 1,
                     etiology_of_shock_choice_valvular_disease == 1 ~ 1,
                     .default = etiology_of_shock_choice_other)) %>%
  select(!etiology_of_shock_choice_post_cardiotomy_shock) %>%
  select(!etiology_of_shock_choice_ards) %>%
  select(!etiology_of_shock_choice_valvular_disease)
 

### filter on age > 18 and femoral cannulation 
ecmo <- ecmo %>%
  mutate(femoral_cannulation = 
           case_when(outflow_cannulation_site == "Femoral Artery" ~ 1,
                     outflow_cannulation_site == "Aorta (Central)" ~ 0,
                     outflow_cannulation_site == "Axillary Artery" ~ 0,
                     v_a_ecmo == "V-A ECMO" ~ 1,
                     .default = NA),
         ischemic_comp = case_when(ischemic_comp_yn == "Yes" ~ 1,
                                   ischemic_comp_yn == "No" ~ 0,
                                   .default = NA)) %>%
  filter(age >= 18 & femoral_cannulation == 1)

# filter for records that have been reviewed and for one with US data
review <- ecmo %>%
  filter(is.na(chart_review_date) == FALSE)
 
us <- ecmo %>%
  filter(us == 1)

```

```{r shock-vars, include=FALSE}
# filter etiology_of_shock_choice

### identify remaining etiology of shock duplicate values
#shock_duplicate <-  us %>%
#  select(c(starts_with("etiology_of_shock_choice"), study_record_id)) %>%
#  pivot_longer(cols = starts_with("etiology_of_shock_choice"), names_to = "etiology_of_shock",
#               names_prefix = "etiology_of_shock_choice_") %>%
#  filter(value == 1) %>%
#  group_by(study_record_id) %>%
#  summarise(n = n()) %>%
#  filter(n > 1)

#us %>%
#  filter(study_record_id %in% shock_duplicate$study_record_id) %>%
#  select(study_record_id, e_mrn, all_of(starts_with("etiology_of_shock"))) %>%
#  View()

us <- us %>%
  select(c(starts_with("etiology_of_shock_choice"), study_record_id)) %>%
  pivot_longer(cols = starts_with("etiology_of_shock_choice"), names_to = "etiology_of_shock",
               names_prefix = "etiology_of_shock_choice_") %>%
  filter(value == 1) %>%   
  right_join(us)


## create a single variable with indication for ecmo
us <- us %>%
  select(c(starts_with("indication_for_ecmo_choice"), study_record_id)) %>%
  pivot_longer(cols = starts_with("indication_for_ecmo_choice"), names_to = "indication_for_ecmo",
               names_prefix = "indication_for_ecmo_choice_") %>%
  filter(value == 1) %>%
  right_join(us)

indication_duplicate <-  us %>%
  select(c(starts_with("indication_for_ecmo_choice"), study_record_id)) %>%
  pivot_longer(cols = starts_with("indication_for_ecmo_choice"), names_to = "indication_for_ecmo",
               names_prefix = "indication_for_ecmo_choice_") %>%
  filter(value == 1) %>%
  group_by(study_record_id) %>%
  summarise(n = n()) %>%
  filter(n > 1)

#ecmo %>%
#  filter(study_record_id %in% shock_duplicate$study_record_id) %>%
#  select(c(starts_with("etiology_of_shock_choice"), study_record_id)) %>%
#  View()


```

There are `r length(unique(ecmo$e_mrn))` subjects within the dataset > 18 with peripheral cannulations. Of which `r length(unique(review$e_mrn))` have been chart reviewed and `r sum(review$us)` had a doppler US assessment of the extremity ipsilateral to the canulation site. With in these patients there were `r nrow(filter(ecmo, amputation_yn == "Yes"))` amputations and `r nrow(filter(ecmo, fasciotomy_yn == "Yes"))` fasciotomies.  

### Characteristics of pateints with ultrasound data

```{r tabl-1-all, echo=FALSE}
review <- review %>%
  mutate(us = as.factor(us))

vars.descriptive <- c("age", "sex", "race_ethnicity", "bmi", "chronic_lung_disease",
                      "DM", "HLD", "hospital_length_of_stay", "icu_los", "pad_yn",
                      "smoking_tobacco_use", "anticoagulation",
                      "shock_after_cardiac_arrest", "cardiogenic_shock", "rosc_prior_to_ecmo",
                      "etiology_of_shock", "mcs_length_of_support",
                      "ecmo_length_of_support", "cannulation_location",
                      "same_side_inflow_outflow", "outflow_cannula_size",
                      "distal_perfusion_cannula_ever",
                      "distal_perfusion_at_cannulation", "ischemic_comp_yn",
                      "amputation_yn", "fasciotomy_yn", "limb_ischemia_yn",
                      "mortality_yn", "ecmo_mortality_yn", "ata_psv", "pta_psv", "peroneal_psv",
                      "peak_ankle_velocity", "log_pav",  "femoral_at_gradient",
                      "femoral_pt_gradient", "max_femoral_pt_at_gradient",
                      "contra_sfa_psv", "mean_sfa_psv",
                      "average_ankle_velocity", "contra_sfa_aav_gradiant",
                      "preop_abi_l", "preop_abi_r",
                      "mean_arterial_pressure_map", "x48h_mean_arterial_pressure",
                      "x72h_mean_arterial_pressure", "vasoactive_inotropic_score",
                      "p_h", "pa_co2", "pa_o2", "lactate",
                      "reperfusion_cath_at_us", "ecmo_flow_at_initiation",
                      "ecmo_flow_at_24h", "ecmo_flow_at_48h", "ecmo_flow_at_72h")
CreateTableOne(vars.descriptive, data = us)

```

### Comparision between pateints with and without ischemic complications in patients with ultrasound data

```{r table-1-ischemic-us, echo=FALSE}
CreateTableOne(vars.descriptive,
              strata = "ischemic_comp_yn",
              data = us)
```

#### Short table

```{r table-1-ischemic-us-short, echo=FALSE}
vars.us.ischemic <- c("etiology_of_shock", "cannulation_location",
                      "outflow_cannula_size",
                      "distal_perfusion_at_cannulation", "same_side_inflow_outflow",
                      "peak_ankle_velocity", "log_pav", "max_femoral_pt_at_gradient",
                      "average_ankle_velocity", "contra_sfa_aav_gradiant")
CreateTableOne(vars.us.ischemic, data = us,
               strata = c("ischemic_comp_yn"),
               argsNonNormal = "outflow_cannula_size")

vars.us.rpc <- c("smoking_tobacco_use", "etiology_of_shock",
                      "ischemic_comp_yn", "log_pav",
                      "peak_ankle_velocity", "max_femoral_pt_at_gradient",
                      "contra_sfa_aav_gradiant", "same_side_inflow_outflow")
CreateTableOne(vars.us.rpc, data = us,
               strata = c("distal_perfusion_at_cannulation"))


CreateTableOne("log_pav", data = filter(us, distal_perfusion_at_cannulation == "Yes"),
               strata = c("ischemic_comp_yn"))

CreateTableOne("log_pav", data = filter(us, same_side_inflow_outflow == "No"),
               strata = c("ischemic_comp_yn"))
CreateTableOne("log_pav", data = filter(us, same_side_inflow_outflow == "Yes"),
               strata = c("ischemic_comp_yn"))

```

#### Comparision between pateints with and without ischemic complications in the entire data set

```{r table-1-ichemic-all, eval=FALSE, include=FALSE}
CreateTableOne(vars.descriptive,
              strata = "ischemic_comp_yn",
              data = ecmo)
              
```

#### Plots comparing variables between pateints with and without ischemic complications

```{r distributions-ischemic, echo=FALSE}

us %>%
  pivot_longer(c(peak_ankle_velocity, max_femoral_pt_at_gradient,
                      average_ankle_velocity, contra_sfa_aav_gradiant),
               names_to = "US_location", values_to = "velocities") %>%
  ggplot(aes(x = velocities )) +
  geom_histogram() +
  facet_wrap(vars(US_location), scales = "free") +
  theme_minimal() 

us %>%
  pivot_longer(c(peak_ankle_velocity, max_femoral_pt_at_gradient,
                      average_ankle_velocity, contra_sfa_aav_gradiant),
               names_to = "US_location", values_to = "velocities") %>%
  ggplot(aes(x = log(velocities + 1))) +
  geom_histogram() +
  facet_wrap(vars(US_location), scales = "free") +
  theme_minimal() 

us %>%
  pivot_longer(peak_ankle_velocity,
               names_to = "US_location", values_to = "velocities") %>%
  ggplot(aes(y = log(velocities + 1), x = as.factor(ischemic_comp_yn),
             fill = as.factor(ischemic_comp_yn))) +
  geom_boxplot() +
  theme_minimal() +
  theme(legend.position = "None") +
  xlab("Ischemic Complication")

us %>%
  ggplot(aes(as.factor(ischemic_comp_yn), outflow_cannula_size)) +
  geom_boxplot() +
  theme_bw() +
  xlab("BMI") +
  theme(legend.position = "none")

us %>%
  ggplot(aes(y = outflow_cannula_size, x = as.factor(ischemic_comp_yn))) +
  geom_boxplot() +
  geom_point(position=position_dodge(width=0.75), aes(group=ischemic_comp_yn)) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  xlab("Ischemic Complication") + 
  scale_fill_discrete(name = "Ischemic Complications",
                      labels = c("No", "Yes"))


us %>%
  pivot_longer(peak_ankle_velocity,
               names_to = "US_location", values_to = "velocities") %>%
  select(ischemic_comp_yn, velocities, US_location,
         distal_perfusion_at_cannulation) %>%
  na.omit() %>%
  ggplot(aes(y = log(velocities + 1), x = as.factor(distal_perfusion_at_cannulation),
             fill = as.factor(ischemic_comp_yn))) +
  geom_boxplot() +
  geom_point(position=position_dodge(width=0.75),aes(group=ischemic_comp_yn)) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  xlab("RP Catheter at Cannulation") + 
  scale_fill_discrete(name = "Ischemic Complications",
                      labels = c("No", "Yes"))

us %>%
  pivot_longer(peak_ankle_velocity,
               names_to = "US_location", values_to = "velocities") %>%
  select(ischemic_comp_yn, velocities, US_location,
         same_side_inflow_outflow) %>%
  na.omit() %>%
  ggplot(aes(y = log(velocities + 1), x = as.factor(same_side_inflow_outflow),
             fill = as.factor(ischemic_comp_yn))) +
  geom_boxplot() +
  geom_point(position=position_dodge(width=0.75),aes(group=ischemic_comp_yn)) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  xlab("Same Sided Arterial and Venous Cannula") + 
  scale_fill_discrete(name = "Ischemic Complications",
                      labels = c("No", "Yes"))


  ggplot(data = us, aes(y = log_pav, x = outflow_cannula_size,
             color = as.factor(ischemic_comp_yn))) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  xlab("Arterial Cannula Size") + 
  scale_color_discrete(name = "Ischemic Complications",
                      labels = c("No", "Yes")) +
    ylab("log(velocities + 1)")


us %>%
  ggplot(aes(bmi, fill = as.factor(ischemic_comp_yn))) +
  geom_histogram(position = "dodge", identity = "stat") +
  facet_grid(ischemic_comp_yn ~ ., scales = "free_y") +
  theme_bw() +
  xlab("BMI") +
  theme(legend.position = "none")
  
us %>%
  pivot_longer(ecmo_flow_at_initiation:ecmo_flow_at_72h,
               names_to = "ecmo_time", values_to = "flow") %>%
  ggplot(aes(x = flow, fill = ischemic_comp_yn)) +
  geom_histogram() +
  facet_grid(rows = vars(ecmo_time), scales = "free_y") +
  theme_minimal() +
  theme(legend.position = "bottom")
  
us %>%
  pivot_longer(c(mean_arterial_pressure_map, x48h_mean_arterial_pressure,
                 x72h_mean_arterial_pressure),
               names_to = "map_time", values_to = "map") %>%
  ggplot(aes(x = map, fill = ischemic_comp_yn)) +
  geom_histogram() +
  facet_grid(rows =  vars(map_time), scales = "free_y") +
  theme_minimal() +
  theme(legend.position = "bottom")

  
```


### Comparision between pateints who survived EMCO and those that did not

```{r table-1-mortality, echo=FALSE}
CreateTableOne(vars.descriptive,
              strata = "ecmo_mortality_yn",
              data = us)
```

#### Short Table

```{r table-1-mortality-short, echo=FALSE}
vars.us.mort <- c("age", "sex","anticoagulation",
                      'peak_ankle_velocity', 'log_pav', 'max_femoral_pt_at_gradient',
                  'contra_sfa_psv', 'mean_sfa_psv',
                      'average_ankle_velocity')
CreateTableOne(vars.us.mort, data = us,
               strata = c("ecmo_mortality_yn"),
               argsNonNormal = c('peak_ankle_velocity', 'max_femoral_pt_at_gradient',
                  'contra_sfa_psv', 'mean_sfa_psv',
                      'average_ankle_velocity'))
```


#### Plots comparing variables between pateints who survived and those that did not

```{r distributions-mortality, echo=FALSE}

us %>%
  pivot_longer(peak_ankle_velocity,
               names_to = "US_location", values_to = "velocities") %>%
  ggplot(aes(y = log(velocities + 1), x = as.factor(ecmo_mortality_yn),
             fill = as.factor(ecmo_mortality_yn))) +
  geom_boxplot() +
  theme_minimal() +
  theme(legend.position = "None") +
  xlab("On EMCO Mortality")

us %>%
  ggplot(aes(age, fill = ecmo_mortality_yn)) +
  geom_histogram(position = "dodge", identity = "stat") +
  facet_grid(mortality_yn ~ ., scales = "free_y") +
  theme_bw() +
  xlab("Age") +
  theme(legend.position = "none")

us %>%
  ggplot(aes(bmi, fill = ecmo_mortality_yn)) +
  geom_histogram(position = "dodge", identity = "stat") +
  facet_grid(mortality_yn ~ ., scales = "free_y") +
  theme_bw() +
  xlab("BMI") +
  theme(legend.position = "none")
  
us %>%
  pivot_longer(c(ata_psv:contra_sfa_psv, mean_sfa_psv,
                 average_ankle_velocity, contra_sfa_aav_gradiant),
               names_to = "US_location", values_to = "velocities") %>%
  ggplot(aes(x = velocities, fill = ecmo_mortality_yn)) +
  geom_density() +
  facet_wrap(vars(US_location), scales = "free_y") +
  theme_minimal() +
  theme(legend.position = "bottom")

us %>%
  ggplot(aes(p_h, fill = ecmo_mortality_yn)) +
  geom_histogram(position = "dodge", identity = "stat") +
  facet_grid(ecmo_mortality_yn ~ ., scales = "free_y") +
  theme_bw() +
  xlab("pH") +
  theme(legend.position = "none")

us %>%
  ggplot(aes(lactate, fill = ecmo_mortality_yn)) +
  geom_histogram(position = "dodge", identity = "stat") +
  facet_grid(ecmo_mortality_yn ~ ., scales = "free_y") +
  theme_bw() +
  xlab("Lactate") +
  theme(legend.position = "none")


us %>%
  pivot_longer(ecmo_flow_at_initiation:ecmo_flow_at_72h,
               names_to = "ecmo_time", values_to = "flow") %>%
  ggplot(aes(x = flow, fill = mortality_yn)) +
  geom_histogram() +
  facet_grid(rows = vars(ecmo_time), scales = "free_y") +
  theme_bw() +
  theme(legend.position = "bottom")
  
us %>%
  pivot_longer(c(mean_arterial_pressure_map, x48h_mean_arterial_pressure,
                 x72h_mean_arterial_pressure),
               names_to = "map_time", values_to = "map") %>%
  ggplot(aes(x = map, fill = mortality_yn)) +
  geom_histogram() +
  facet_grid(rows =  vars(map_time), scales = "free_y") +
  theme_bw() +
  theme(legend.position = "bottom")
```
## Regression Analysis and Prediction

### Prediction of Ischemic events while on ECMO

#### Univariate Regression

```{r univar-logistic-regression-ischemia, echo=FALSE}
# mortality univariate
uni_models <- vars.descriptive %>%                           # begin with variables of interest
  str_c("ischemic_comp_yn ~ ", .) %>%  # combine each variable into formula ("outcome ~ variable of interest")
  
  # iterate through each univariate formula
  map(
    .f = ~glm(                            # pass the formulas one-by-one to glm()
    formula = as.formula(.x),             # within glm(), the string formula is .x
    family = "binomial",                  # specify type of glm (logistic)
                data = us)) %>%    # dataset
# tidy up each of the glm regression outputs from above
  map(.f = ~tidy(.x,
                 exponentiate = TRUE,
                 conf.int = TRUE)) %>%
 # collapse the list of regression outputs in to one data frame
  bind_rows() %>%
# round all numeric columns
  mutate(across(where(is.numeric), round, digits = 5))
    
uni_models %>%
    filter(term != "(Intercept)" & p.value < 0.2) %>%
  print(n = nrow(uni_models))

#correlation_matrix <- rcorr(as.matrix(uni_var_vqi_data_num), type = "pearson")
#view(correlation_matrix)
#ggpairs(uni_var_vqi_data_num)

data_cov_uni <- c("race_ethnicity", "same_side_inflow_outflow", "outflow_cannula_size",
                  "log_pav", "distal_perfusion_at_cannulation")

univ_tab <- us %>% 
  dplyr::select(all_of(data_cov_uni), outcome = ischemic_comp_yn) %>% ## select variables of interest

  tbl_uvregression(                         ## produce univariate table
    method = glm,                           ## define regression want to run (generalised linear model)
    y = outcome,                            ## define outcome variable
    method.args = list(family = binomial),  ## define what type of glm want to run (logistic)
    exponentiate = TRUE                     ## exponentiate to produce odds ratios (rather than log odds)
  )

## view univariate results table 
univ_tab


```
#### Multivariate Logistic Regression

```{r multivar-logistic-regression-ischemia, echo=FALSE}

## run a regression with all variables of interest 
mv_reg <- data_cov_uni %>%  ## begin with vector of explanatory column names
  str_c(collapse = "+") %>%     ## combine all names of the variables of interest separated by a plus
  str_c("ischemic_comp_yn ~ ", .) %>%    ## combine the names of variables of interest with outcome in formula style
  glm(family = "binomial",      ## define type of glm as logistic,
      data = us)          ## define your dataset


model_ischemic_full <- 
  glm(ischemic_comp_yn ~ race_ethnicity + same_side_inflow_outflow +
        outflow_cannula_size + log_pav + distal_perfusion_at_cannulation,
      data = us)
summary(model_ischemic_full)

model_ischemic_step <- 
  glm(ischemic_comp_yn ~ race_ethnicity + same_side_inflow_outflow +
        outflow_cannula_size + log_pav + distal_perfusion_at_cannulation,
      data = us,
      na.action = "na.exclude")
summary(model_ischemic_step)


confusionMatrix(as.factor(ifelse(fitted(model_ischemic_step) > .5, "Yes", "No")),
                reference = as.factor(ifelse(us$ischemic_comp_yn == 1, "Yes", "No")),
                positive = "Yes")

invisible(plot(roc(model_ischemic_step$y,
                   model_ischemic_step$fitted.values),
               print.auc = T, 
               col = "blue"))

library(arm)
#likelihood ratio test for different models as needed
#lrtest()

#This binned plot looks bad
binnedplot(na.omit(fitted(model_ischemic_step)), 
          na.omit(residuals(model_ischemic_step, type = "response")), 
           nclass = NULL, 
           xlab = "Expected Values", 
           ylab = "Average residual", 
           main = "Binned residual plot", 
           cex.pts = 0.8, 
           col.pts = 1, 
          col.int = "gray")

detach("package:arm", unload = TRUE)
detach("package:MASS", unload = TRUE)

mv_tab_base <- model_ischemic_step %>% 
  broom::tidy(exponentiate = TRUE, conf.int = TRUE) %>%  ## get a tidy dataframe of estimates 
  mutate(across(where(is.numeric), round, digits = 4))

mv_tab <- tbl_regression(model_ischemic_step, exponentiate = TRUE)
mv_tab

tbl_merge(
  tbls = list(univ_tab, mv_tab),                          # combine
  tab_spanner = c("**Univariate**", "**Multivariable**")) # set header names

```

#### ROC Curve for Log Peak Ankle Velocity to Predict Ischemia
```{r roc-logpav-ischemia, echo=FALSE}

pROC_obj <- us %>%
  roc_(as.factor("ischemic_comp_yn"),'log_pav' ,
       smoothed = TRUE,
            # arguments for ci
            ci=TRUE, ci.alpha=0.9, stratified=FALSE,
            # arguments for plot
            plot=TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, grid=TRUE,
            print.auc=TRUE, show.thres=TRUE)

```

### Prediction of Mortality while on ECMO

#### Univariate Regression

```{r univar-logistic-regression-mortality, echo=FALSE}
# mortality univariate
uni_models <- vars.descriptive %>%                           # begin with variables of interest
  str_c("ecmo_mortality_yn ~ ", .) %>%  # combine each variable into formula ("outcome ~ variable of interest")
  
  # iterate through each univariate formula
  map(
    .f = ~glm(                            # pass the formulas one-by-one to glm()
    formula = as.formula(.x),             # within glm(), the string formula is .x
    family = "binomial",                  # specify type of glm (logistic)
                data = us)) %>%    # dataset
# tidy up each of the glm regression outputs from above
  map(.f = ~tidy(.x,
                 exponentiate = TRUE,
                 conf.int = TRUE)) %>%
 # collapse the list of regression outputs in to one data frame
  bind_rows() %>%
# round all numeric columns
  mutate(across(where(is.numeric), round, digits = 5))
    
uni_models %>%
    filter(term != "(Intercept)" & p.value < 0.1) %>%
  print(n = nrow(uni_models))

#correlation_matrix <- rcorr(as.matrix(uni_var_vqi_data_num), type = "pearson")
#view(correlation_matrix)
#ggpairs(uni_var_vqi_data_num)

data_cov_uni <- c("age", "sex", "anticoagulation",
                  "log_pav",
                  "max_femoral_pt_at_gradient")

univ_tab <- us %>% 
  dplyr::select(all_of(data_cov_uni), outcome = ecmo_mortality_yn) %>% ## select variables of interest

  tbl_uvregression(                         ## produce univariate table
    method = glm,                           ## define regression want to run (generalised linear model)
    y = outcome,                            ## define outcome variable
    method.args = list(family = binomial),  ## define what type of glm want to run (logistic)
    exponentiate = TRUE                     ## exponentiate to produce odds ratios (rather than log odds)
  )

## view univariate results table 
univ_tab


```
#### Multivariate Logistic Regression

```{r multivar-logistic-regression-mortality, echo=FALSE}

## run a regression with all variables of interest 
mv_reg <- data_cov_uni %>%  ## begin with vector of explanatory column names
  str_c(collapse = "+") %>%     ## combine all names of the variables of interest separated by a plus
  str_c("ecmo_mortality_yn ~ ", .) %>%    ## combine the names of variables of interest with outcome in formula style
  glm(family = "binomial",      ## define type of glm as logistic,
      data = us)          ## define your dataset


model_mort_full <- 
  glm(ecmo_mortality_yn ~ age + sex + anticoagulation +
        log_pav + max_femoral_pt_at_gradient +
        ecmo_flow_at_72h, data = us)
summary(model_mort_full)

model_mort_step <- 
  glm(ecmo_mortality_yn ~ age +
        log_pav, data = us,
      na.action = "na.exclude")
summary(model_mort_step)


confusionMatrix(as.factor(ifelse(fitted(model_mort_step) > .1, "Yes", "No")),
                reference = as.factor(ifelse(us$ecmo_mortality_yn == 1, "Yes", "No")),
                positive = "Yes")

invisible(plot(roc(model_mort_step$y,
                   model_mort_step$fitted.values),
               print.auc = T, 
               col = "blue"))

library(arm)
#likelihood ratio test for different models as needed
#lrtest()

#This binned plot looks bad
binnedplot(na.omit(fitted(model_mort_step)), 
          na.omit(residuals(model_mort_step, type = "response")), 
           nclass = NULL, 
           xlab = "Expected Values", 
           ylab = "Average residual", 
           main = "Binned residual plot", 
           cex.pts = 0.8, 
           col.pts = 1, 
          col.int = "gray")

detach("package:arm", unload = TRUE)
detach("package:MASS", unload = TRUE)

mv_tab_base <- model_mort_step %>% 
  broom::tidy(exponentiate = TRUE, conf.int = TRUE) %>%  ## get a tidy dataframe of estimates 
  mutate(across(where(is.numeric), round, digits = 4))

mv_tab <- tbl_regression(model_mort_step, exponentiate = TRUE)
mv_tab

tbl_merge(
  tbls = list(univ_tab, mv_tab),                          # combine
  tab_spanner = c("**Univariate**", "**Multivariable**")) # set header names

```

#### ROC Curve for Log Peak Ankle Velocity to predict ECMO Mortality
```{r roc-logpav-mort, echo=FALSE}

pROC_obj <- us %>%
  roc_(as.factor("ecmo_mortality_yn"),'log_pav' ,
       smoothed = TRUE,
            # arguments for ci
            ci=TRUE, ci.alpha=0.9, stratified=FALSE,
            # arguments for plot
            plot=TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, grid=TRUE,
            print.auc=TRUE, show.thres=TRUE)

```


```{r eval=FALSE, include=FALSE}

sas_vars <- read.csv("VascularECMOdataset_sas_vars.csv", header = TRUE)

ecmo.bst <- us %>%
  select(all_of(names(sas_vars)))
write_csv(ecmo.bst, "VascularECMOdataset_bst_v2.csv")

```


