---
title: "ECMO US"
author: "Joshua Geiger"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(broom)
library(janitor)
library(readxl)
library(tableone)
library(ROCR)
library(stringr)

ecmo <- read_xlsx("raw_data/VascularECMOdataset_11_5_23_mrn_cleaned.xlsx",
                 na = c("", "NA", ".", "9999", "Not applicable")) %>%
  clean_names()

```

A total of `r length(unique(ecmo$e_mrn))` patients were identified since January 2013 with a total of `r nrow(ecmo)` ecmo cannulations. The data set was filtered for patients > 18 years old who had peripheral VA - ECMO (Femoral cannulation).

## Descriptive statistics

```{r data-cleaning, include=FALSE}

ecmo <- ecmo %>%
  mutate(sex = case_when(sex == "F" ~ "Female",
                         sex == "M" ~ "Male",
                         .default = sex),
####demographics          
         age = as.numeric(age),
        race_ethnicity = case_when(race_ethnicity == "Asian" ~ "Other",
                                   race_ethnicity == "Hispanic/Latino" ~ "Other",
                                   race_ethnicity ==  "Native American or Alaska Native" ~ "Other",
                                   .default = race_ethnicity),
        cannulation_location = case_when(cannulation_location == "Non-ICU WARD" ~ "Other",
                                         cannulation_location == "OSH Transfer, already on ECMO" ~ "OSH",
                                         cannulation_location == "OSH (transfer)" ~ "OSH",
                                         cannulation_location == "Team goes to OSH ‚Üí cannulates patient ‚Üí transfers to Primary Facility" ~ "OSH",
                                         .default = cannulation_location),
#### comorbidities
        chronic_lung_disease = case_when(past_medical_history_choice_chronic_lung_disease == 1 ~ "Yes", 
                                         past_medical_history_choice_chronic_lung_disease == 0 ~ "No",
                         .default = NA),
        DM = case_when(past_medical_history_choice_diabetes_mellitus == 1 ~ "Yes",
                       past_medical_history_choice_diabetes_mellitus == 0 ~ "No",
                       .default = NA),
        HLD = case_when(past_medical_history_choice_hyperlipidemia == 1 ~ "Yes",
                       past_medical_history_choice_hyperlipidemia == 0 ~ "No",
                       .default = NA),
        shock_after_cardiac_arrest = case_when(etiology_of_shock_choice_cardiac_arrest_code == 1 ~ "Yes",
                                etiology_of_shock_choice_cardiac_arrest_code == 0 ~ "No",
                                .default = NA),
        anticoagulation = case_when(type_of_anticoagulation_choice_bivalirudin_drip == 1 ~ "Yes",
                                    type_of_anticoagulation_choice_heparin_drip == 1 ~ "Yes",
                                    type_of_anticoagulation_choice_lovenox_injections == 1 ~ "Yes",
                                    type_of_anticoagulation_choice_bivalirudin_drip == 0 &
                                      type_of_anticoagulation_choice_heparin_drip == 0 &
                                      type_of_anticoagulation_choice_lovenox_injections == 0 ~ "No",
                                    .default = NA),
#### US velocities
         sfa_distal_psv = case_when(!is.na(sfa_distal_pvs_on_ecmo_pre_rpc) ~ sfa_distal_pvs_on_ecmo_pre_rpc,
                                    !is.na(sfa_distal_pvs_on_ecmo_post_rpc) ~ sfa_distal_pvs_on_ecmo_post_rpc,
                                    .default = NA),
         sfa_mid_psv = case_when(!is.na(sfa_mid_pvs_on_ecmo_pre_rpc) ~ sfa_mid_pvs_on_ecmo_pre_rpc,
                                    !is.na(sfa_mid_pvs_on_ecmo_post_rpc) ~ sfa_mid_pvs_on_ecmo_post_rpc,
                                 .default = NA),
         sfa_prox_psv = case_when(!is.na(sfa_prox_pvs_on_ecmo_pre_rpc) ~ sfa_prox_pvs_on_ecmo_pre_rpc,
                                    !is.na(sfa_prox_pvs_on_ecmo_post_rpc) ~ sfa_prox_pvs_on_ecmo_post_rpc,
                                  .default = NA),
         sfa_prox_mid_psv = case_when(!is.na(sfa_prox_pvs_on_ecmo_pre_rpc) ~ sfa_prox_pvs_on_ecmo_pre_rpc,
                             !is.na(sfa_mid_pvs_on_ecmo_pre_rpc) ~ sfa_mid_pvs_on_ecmo_pre_rpc,
                             !is.na(sfa_prox_pvs_on_ecmo_post_rpc) ~ sfa_prox_pvs_on_ecmo_post_rpc,
                             !is.na(sfa_mid_pvs_on_ecmo_post_rpc) ~ sfa_mid_pvs_on_ecmo_post_rpc,
                             .default = NA),
         ata_psv = case_when(!is.na(ata_pvs_on_ecmo_pre_rpc) ~ ata_pvs_on_ecmo_pre_rpc,
                                    !is.na(ata_pvs_on_ecmo_post_rpc) ~ ata_pvs_on_ecmo_post_rpc),
         pta_psv = case_when(!is.na(pta_pvs_on_ecmo_pre_rpc) ~ pta_pvs_on_ecmo_pre_rpc,
                                    !is.na(pta_pvs_on_ecmo_post_rpc) ~ pta_pvs_on_ecmo_post_rpc),
         peroneal_psv = case_when(!is.na(peroneal_pvs_on_ecmo_pre_rpc) ~ peroneal_pvs_on_ecmo_pre_rpc,
                                    !is.na(peroneal_pvs_on_ecmo_post_rpc) ~ peroneal_pvs_on_ecmo_post_rpc),
         peak_ankle_velocity = pmax(ata_psv, pta_psv, peroneal_psv, na.rm = TRUE),
         femoral_at_gradient = sfa_prox_mid_psv - ata_psv,
         femoral_pt_gradient = sfa_prox_mid_psv - pta_psv,
         contra_sfa_psv = case_when(!is.na(contra_sfa_prox_pvs_on_ecmo_pre_rpc) ~ contra_sfa_prox_pvs_on_ecmo_pre_rpc,
                                    !is.na(contra_sfa_prox_psv_on_ecmo_post_rpc) ~ contra_sfa_prox_psv_on_ecmo_post_rpc,
                                  .default = NA),
         preop_abi_l = case_when(preop_abi_l == "Non-compressible" ~ "2.00",
                                 .default = preop_abi_l),
         preop_abi_l = as.numeric(preop_abi_l),
         preop_abi_r = case_when(preop_abi_r == "Non-compressible" ~ "2.00",
                                 .default = preop_abi_r),
         preop_abi_r = as.numeric(preop_abi_r),
         preop_abi_r = case_when(preop_abi_r >2.0 ~ 2.0,
                                 .default = preop_abi_r),
         preop_abi_l = case_when(preop_abi_l >2.0 ~ 2.0,
                                 .default = preop_abi_l),
 
#### outcomes
         hospital_length_of_stay = as.numeric(hospital_length_of_stay),
         amputation_yn = case_when(amputation == 1 ~ "Yes",
                                   amputation == 0 ~ "No",
                                   .default = NA),
         fasciotomy_yn = case_when(fasciotomy == 1 ~ "Yes",
                                   fasciotomy == 0 ~ "No",
                                   .default = NA),
         mortality_yn = case_when(mortality == 1 ~ "Yes",
                                   mortality == 0 ~ "No",
                                   .default = NA),
         limb_ischemia_yn = case_when(hypoperfusion_limb_ischemia == 1 ~ "Yes",
                                      fasciotomy == 1 ~ "Yes",
                                      hypoperfusion_limb_ischemia == 0 ~ "No",
                                      fasciotomy == 0 ~ "No",
                                     .default = NA),
         ischemic_comp_yn = case_when(amputation_yn == "Yes" | fasciotomy_yn == "Yes" | limb_ischemia_yn == "Yes" ~ "Yes",
                                     amputation_yn == "No" ~ "No",
                                     fasciotomy_yn == "No" ~ "No",
                                     limb_ischemia_yn == "No" ~ "No",
                                     .default = NA)) %>%
  ### more comorbidities
        mutate(pad_yn = case_when(pad == 1 ~ "Yes",
                           preop_abi_l < 0.9 ~ "Yes",
                           preop_abi_r < 0.9 ~ "Yes",
                           pad == 0 ~ "No",
                           preop_abi_l > 0.9 & preop_abi_l < 1.2 &
                             preop_abi_r > 0.9 & preop_abi_l < 1.2 ~ "No",
                           .default = NA),
          distal_perfusion_cannula_ever = case_when(distal_perfusion_cannula == "Yes" ~ "Yes",
                                                      distal_perfusion_cannula == "No" ~ "No",
                                                      .default = NA),
          distal_perfusion_at_cannulation =
            case_when(distal_perfusion_cannula_done_at_the_time_of_initial_cannulation == "Yes" ~ "Yes",
                      distal_perfusion_cannula_done_at_the_time_of_initial_cannulation == "No" ~ "No",
                      .default = NA),
            same_side_inflow_outflow = case_when(outflow_cannulation_side == inflow_cannulation_side ~ "Yes",
                                               outflow_cannulation_side != inflow_cannulation_side ~ "No",
                                               .default = NA)) %>%
  select(!etiology_of_shock_choice_cardiac_arrest_code) %>%
  rename(rosc_prior_to_ecmo = return_of_spontaneous_circulation_rosc_prior_to_ecmo_deployment) %>%
  mutate(us = case_when(sfa_distal_pvs_on_ecmo_pre_rpc >= 0 ~ 1,
                        sfa_mid_pvs_on_ecmo_pre_rpc >= 0 ~ 1,
                        sfa_prox_pvs_on_ecmo_pre_rpc  >= 0 ~ 1,
                        pop_pvs_on_ecmo_pre_rpc >= 0 ~ 1,
                        peroneal_pvs_on_ecmo_pre_rpc >= 0 ~ 1,
                        ata_pvs_on_ecmo_pre_rpc >= 0 ~ 1,
                        pta_pvs_on_ecmo_pre_rpc >= 0 ~ 1,
                        sfa_prox_pvs_on_ecmo_post_rpc >= 0 ~ 1,
                        sfa_mid_pvs_on_ecmo_post_rpc >= 0 ~ 1,
                        sfa_distal_pvs_on_ecmo_post_rpc >= 0 ~ 1,
                        pop_pvs_on_ecmo_post_rpc >= 0 ~ 1,
                        peroneal_pvs_on_ecmo_post_rpc >= 0 ~ 1,
                        ata_pvs_on_ecmo_post_rpc >= 0 ~ 1,
                        pta_pvs_on_ecmo_post_rpc >= 0 ~ 1,
                          .default = 0),
         reperfusion_cath_at_us = case_when(!is.na(sfa_distal_pvs_on_ecmo_pre_rpc) ~ 1,
                        !is.na(sfa_mid_pvs_on_ecmo_pre_rpc) ~ 1,
                        !is.na(sfa_prox_pvs_on_ecmo_pre_rpc) ~ 1,
                        !is.na(pop_pvs_on_ecmo_pre_rpc) ~ 1,
                        !is.na(peroneal_pvs_on_ecmo_pre_rpc) ~ 1,
                        !is.na(ata_pvs_on_ecmo_pre_rpc) ~ 1,
                        !is.na(pta_pvs_on_ecmo_pre_rpc) ~ 1,
                          .default = 0)) %>%
### calculate means across columns for velocities  
  rowwise() %>%
  mutate(mean_sfa_psv = mean(c_across(c(sfa_prox_psv, sfa_mid_psv, sfa_distal_psv)), na.rm = TRUE),
         average_ankle_velocity = mean(c_across(c(ata_psv, pta_psv, peroneal_psv)), na.rm = TRUE)) 

### variable dependent on means
ecmo <- ecmo %>%
  mutate(contra_sfa_aav_gradiant = contra_sfa_psv - average_ankle_velocity)

### identify etiology of shock duplicate values
shock_duplicate <-  ecmo %>%
  select(c(starts_with("etiology_of_shock_choice"), study_record_id)) %>%
  pivot_longer(cols = starts_with("etiology_of_shock_choice"), names_to = "etiology_of_shock",
               names_prefix = "etiology_of_shock_choice_") %>%
  filter(value == 1) %>%
  group_by(study_record_id) %>%
  summarise(n = n()) %>%
  filter(n > 1)

# remove "other" duplicates
ecmo <- ecmo %>%
  mutate(etiology_of_shock_choice_other = 
           case_when(study_record_id %in% shock_duplicate$study_record_id ~ 0,
                     etiology_of_shock_choice_post_cardiotomy_shock == 1 ~ 1,
                     etiology_of_shock_choice_ards == 1 ~ 1,
                     etiology_of_shock_choice_valvular_disease == 1 ~ 1,
                     .default = etiology_of_shock_choice_other)) %>%
  select(!etiology_of_shock_choice_post_cardiotomy_shock) %>%
  select(!etiology_of_shock_choice_ards) %>%
  select(!etiology_of_shock_choice_valvular_disease)
 
### identify remaining etiology of shock duplicate values
shock_duplicate <-  ecmo %>%
  select(c(starts_with("etiology_of_shock_choice"), study_record_id)) %>%
  pivot_longer(cols = starts_with("etiology_of_shock_choice"), names_to = "etiology_of_shock",
               names_prefix = "etiology_of_shock_choice_") %>%
  filter(value == 1) %>%
  group_by(study_record_id) %>%
  summarise(n = n()) %>%
  filter(n > 1)


### filter on age > 18 and femoral cannulation 
ecmo <- ecmo %>%
  mutate(femoral_cannulation = 
           case_when(outflow_cannulation_site == "Femoral Artery" ~ 1,
                     outflow_cannulation_site == "Aorta (Central)" ~ 0,
                     outflow_cannulation_site == "Axillary Artery" ~ 0,
                     v_a_ecmo == "V-A ECMO" ~ 1,
                     .default = NA),
         ischemic_comp = case_when(ischemic_comp_yn == "Yes" ~ 1,
                                   ischemic_comp_yn == "No" ~ 0,
                                   .default = NA)) %>%
  filter(age >= 18 & femoral_cannulation == 1)


#ecmo %>%
#  filter(study_record_id %in% shock_duplicate$study_record_id) %>%
#  select(c(starts_with("etiology_of_shock_choice"), study_record_id)) %>%
#  View()
  
ecmo <- ecmo %>%
  select(c(starts_with("etiology_of_shock_choice"), study_record_id)) %>%
  pivot_longer(cols = starts_with("etiology_of_shock_choice"), names_to = "etiology_of_shock",
               names_prefix = "etiology_of_shock_choice_") %>%
  filter(value == 1) %>%
  right_join(ecmo)


## create a single variable with indication for ecmo
ecmo <- ecmo %>%
  select(c(starts_with("indication_for_ecmo_choice"), study_record_id)) %>%
  pivot_longer(cols = starts_with("indication_for_ecmo_choice"), names_to = "indication_for_ecmo",
               names_prefix = "indication_for_ecmo_choice_") %>%
  filter(value == 1) %>%
  right_join(ecmo)

# filter for records that have been reviewed and for one with US data
review <- ecmo %>%
  filter(is.na(chart_review_date) == FALSE)
 

us <- ecmo %>%
  filter(us == 1)



```

There are `r length(unique(ecmo$e_mrn))` subjects within the dataset > 18 with peripheral cannulations. Of which `r length(unique(review$e_mrn))` have been chart reviewed and `r sum(review$us)` had a doppler US assessment of the extremity ipsilateral to the canulation site. With in these patients there were `r nrow(filter(ecmo, amputation_yn == "Yes"))` amputations and `r nrow(filter(ecmo, fasciotomy_yn == "Yes"))` fasciotomies.  

#### Characteristics of pateints with ultrasound data

```{r tabl-1-all, echo=FALSE}
review <- review %>%
  mutate(us = as.factor(us))

vars.descriptive <- c("age", "sex", "race_ethnicity", "bmi", "chronic_lung_disease",
                      "DM", "HLD", "hospital_length_of_stay", "icu_los", "pad_yn",
                      "smoking_tobacco_use", "anticoagulation",
                      "shock_after_cardiac_arrest", "rosc_prior_to_ecmo",
                      "etiology_of_shock", "mcs_length_of_support",
                      "ecmo_length_of_support", "cannulation_location",
                      "same_side_inflow_outflow", "outflow_cannula_size",
                      "distal_perfusion_cannula_ever",
                      "distal_perfusion_at_cannulation", "ischemic_comp_yn",
                      "amputation_yn", "fasciotomy_yn", "limb_ischemia_yn",
                      "mortality_yn", "ata_psv", "pta_psv", "peroneal_psv",
                      "peak_ankle_velocity", "femoral_at_gradient",
                      "femoral_pt_gradient", "contra_sfa_psv", "mean_sfa_psv",
                      "average_ankle_velocity", "contra_sfa_aav_gradiant",
                      "preop_abi_l", "preop_abi_r",
                      "mean_arterial_pressure_map", "x48h_mean_arterial_pressure",
                      "x72h_mean_arterial_pressure", "vasoactive_inotropic_score",
                      "p_h", "pa_co2", "pa_o2", "lactate",
                      "reperfusion_cath_at_us", "ecmo_flow_at_initiation",
                      "ecmo_flow_at_24h", "ecmo_flow_at_48h", "ecmo_flow_at_72h")
CreateTableOne(vars.descriptive, data = us)

```
#### Comparision between pateints with and without ischemic complications in patients with ultrasound data

```{r table-1-ichemic-us}
CreateTableOne(vars.descriptive,
              strata = "ischemic_comp_yn",
              data = us)
```
#### Comparision between pateints with and without ischemic complications in the entire data set

```{r table-1-ichemic-all}
CreateTableOne(vars.descriptive,
              strata = "ischemic_comp_yn",
              data = ecmo)
              
```
#### Plots comparing variables between pateints with and without ischemic complications

```{r distributions-ischemic, echo=FALSE}

us %>%
  pivot_longer(c(ata_psv:contra_sfa_psv, mean_sfa_psv,
                 average_ankle_velocity, contra_sfa_aav_gradiant),
               names_to = "US_location", values_to = "velocities") %>%
  ggplot(aes(x = velocities, fill = ischemic_comp_yn)) +
  geom_histogram() +
  facet_wrap(vars(US_location), scales = "free_y") +
  theme_minimal() +
  theme(legend.position = "bottom")

ecmo %>%
  ggplot(aes(bmi, fill = ischemic_comp_yn)) +
  geom_histogram(position = "dodge", identity = "stat") +
  facet_grid(ischemic_comp_yn ~ ., scales = "free_y") +
  theme_bw() +
  xlab("BMI") +
  theme(legend.position = "none")
  
ecmo %>%
  pivot_longer(ecmo_flow_at_initiation:ecmo_flow_at_72h,
               names_to = "ecmo_time", values_to = "flow") %>%
  ggplot(aes(x = flow, fill = ischemic_comp_yn)) +
  geom_histogram() +
  facet_grid(rows = vars(ecmo_time), scales = "free_y") +
  theme_minimal() +
  theme(legend.position = "bottom")
  
ecmo %>%
  pivot_longer(c(mean_arterial_pressure_map, x48h_mean_arterial_pressure,
                 x72h_mean_arterial_pressure),
               names_to = "map_time", values_to = "map") %>%
  ggplot(aes(x = map, fill = ischemic_comp_yn)) +
  geom_histogram() +
  facet_grid(rows =  vars(map_time), scales = "free_y") +
  theme_minimal() +
  theme(legend.position = "bottom")

  
```
#### Comparision between pateints who survived and those that did not in the whole data set

```{r table-1-mortality}
CreateTableOne(vars.descriptive,
              strata = "mortality_yn",
              data = us)
```

#### Plots comparing variables between pateints who survived and those that did not

```{r distributions-mortality, echo=FALSE}

ecmo %>%
  ggplot(aes(age, fill = mortality_yn)) +
  geom_histogram(position = "dodge", identity = "stat") +
  facet_grid(mortality_yn ~ ., scales = "free_y") +
  theme_bw() +
  xlab("Age") +
  theme(legend.position = "none")

ecmo %>%
  ggplot(aes(bmi, fill = mortality_yn)) +
  geom_histogram(position = "dodge", identity = "stat") +
  facet_grid(mortality_yn ~ ., scales = "free_y") +
  theme_bw() +
  xlab("BMI") +
  theme(legend.position = "none")
  
us %>%
  pivot_longer(c(ata_psv:contra_sfa_psv, mean_sfa_psv,
                 average_ankle_velocity, contra_sfa_aav_gradiant),
               names_to = "US_location", values_to = "velocities") %>%
  ggplot(aes(x = velocities, fill = mortality_yn)) +
  geom_density() +
  facet_wrap(vars(US_location), scales = "free_y") +
  theme_minimal() +
  theme(legend.position = "bottom")

ecmo %>%
  ggplot(aes(p_h, fill = mortality_yn)) +
  geom_histogram(position = "dodge", identity = "stat") +
  facet_grid(mortality_yn ~ ., scales = "free_y") +
  theme_bw() +
  xlab("pH") +
  theme(legend.position = "none")

ecmo %>%
  ggplot(aes(lactate, fill = mortality_yn)) +
  geom_histogram(position = "dodge", identity = "stat") +
  facet_grid(mortality_yn ~ ., scales = "free_y") +
  theme_bw() +
  xlab("Lactate") +
  theme(legend.position = "none")


ecmo %>%
  pivot_longer(ecmo_flow_at_initiation:ecmo_flow_at_72h,
               names_to = "ecmo_time", values_to = "flow") %>%
  ggplot(aes(x = flow, fill = mortality_yn)) +
  geom_histogram() +
  facet_grid(rows = vars(ecmo_time), scales = "free_y") +
  theme_bw() +
  theme(legend.position = "bottom")
  
ecmo %>%
  pivot_longer(c(mean_arterial_pressure_map, x48h_mean_arterial_pressure,
                 x72h_mean_arterial_pressure),
               names_to = "map_time", values_to = "map") %>%
  ggplot(aes(x = map, fill = mortality_yn)) +
  geom_histogram() +
  facet_grid(rows =  vars(map_time), scales = "free_y") +
  theme_bw() +
  theme(legend.position = "bottom")
```


```{r eval=FALSE, include=FALSE}

ecmo.bst <- us %>%
  select(all_of(vars.descriptive))
write_csv(ecmo.bst, "VascularECMOdataset_bst.csv")

# Fit the logistic regression model
fit <- glm(is.numeric(ischemic_comp_yn) ~ sfa_distal_pvs, data = ecmo, family = binomial)

# Generate predictions
predictions <- predict(fit, type = "response")

# Create a prediction object
pred <- prediction(predict(fit), labels)

# Create a performance object
perf <- performance(pred, "tpr", "fpr")

# Plot the ROC curve
plot(perf)

```

